Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-25 _tex/paper2.tex
Built site for gh-pages

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on the validity of CJ scores indicates their
capacity to represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Moreover, research on CJ's practical applicability highlights its
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. The study then addresses these
issues by extending Thurstone's general form through a systematic and
integrated approach based on causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's theory to
address these challenges. The extension integrates core theoretical
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the implications of the findings and explores avenues for
future research. Finally, Section~\ref{sec-conclusion} summarizes the
study's key insights.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Furthermore, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

@@

\section{Two prominent issues in CJ literature}\label{sec-theory-issues}

This section identifies and discusses the two prominent issues that can
undermine the reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

\subsection{The Case V and the statistical analysis of CJ
data}\label{sec-theory-issue1}

Thurstone noted from the outset that the general form of the theory, as
outlined in Section~\ref{sec-thurstone_theory}, gave rise to a trait
scaling problem. Specifically, the model required estimating more
``unknown'' parameters than the available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
the practical implementation of the theory, he developed five cases
derived from this general form, each progressively incorporating
additional simplifying assumptions into the model.

In Case I, Thurstone postulated that pairs of stimuli would maintain a
constant correlation across all comparisons. In Case II, he allowed
multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

Notably, despite relying on the most extensive set of simplifying
assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli's discriminal processes,
the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}. Furthermore,
he acknowledged that some assumptions could prove problematic when
researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice, namely the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this concept. In this experiment, the
researcher observes the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\).
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more narrow compared to situations where their
dispersions differ. Consequently, the discriminal difference is more
likely to favor Text B over Text A (shaded gray area)

@@

\caption{\label{fig-casev_issues}The effect of dispersion discrepancies
and stimulus correlation on the distribution of the discriminal
difference.}

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between the stimuli
and their respective discriminal processes
\citep[pp.~373]{Thurstone_1927a}. Therefore, the outcome's ability to
reflect the ``true'' differences between stimuli largely depends on the
validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
particularly the assumption of equal dispersions. For instance, when
researchers observe a sample of outcomes favoring Text B over Text A and
correctly assume equal dispersions between the texts, the BTL model
estimates a discriminal difference distribution that accurately
represents the ``true'' discriminal difference of the texts. This
scenario is illustrated in Figure~\ref{fig-dispersion}, when the model's
discriminal difference distribution aligns with the ``true''
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of the discriminal difference
then ensures reliable estimates for the texts' discriminal processes
{(citation needed?)}.

However, Thurstone argued that the assumption of equal dispersions may
not be applicable when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of this violation may already be present in the CJ
literature in the form of misfit statistics, which measure judgment
discrepancies associated with specific stimuli
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For example, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than those
involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
These discrepancies, in turn, suggest that the discriminal differences
for ``misfit'' texts have broader distributions, indicating that their
discriminal processes may also exhibit more variation than that of other
texts. A similar line of reasoning applies to the concept of ``misfit''
judges, whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of the stimuli or the judges
themselves. These ``misfit'' judges and their associated deviations can
give rise to additional statistical and measurement issues, which we
discuss in more detail in Section~\ref{sec-theory-issue1b}.

Thus, erroneously assuming equal dispersions between stimuli, can give
rise to significant statistical and measurement issues. For instance,
the model may overestimate the degree to which the outcome accurately
reflects the ``true'' discriminal differences between stimuli. This
overestimation can result in researchers drawing spurious conclusions
about these differences \citep[pp.~370]{McElreath_2020} and, by
extension, about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this issue when the model's
discriminal difference distribution aligns with the thick continuous
line for \(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
that misfit statistics highlight these critical differences in
dispersions, the conventional CJ practice of excluding stimuli based on
these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on which stimuli are excluded from the analysis.

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
similar thought experiment, as the one in
Section~\ref{sec-theory-issue1a}, can illustrate this concept. The
experiment now assumes that the discriminal dispersions for both texts
remain constant. Figure~\ref{fig-correlation} reveals that as the
correlation between the texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing affects the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
outcome. Furthermore, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less narrow compared to scenarios where the texts are highly
correlated. Consequently, the discriminal difference is less likely to
favor Text B over Text A (shaded gray area).

Again, in experimental practice, researchers approach this process in
reverse. They begin by observing the sample of outcomes favoring Text B
over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is also illustrated
in Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the accuracy of the discriminal difference ensures reliable
estimates for the discriminal processes of the texts {(citation
needed?)}.

Notably, Thurstone attributed the lack of correlation between stimuli to
the cancellation of potential judges' biases. He argued that this
cancellation resulted from two opposing and equally weighted effects
occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
\citet{Andrich_1978} provided a mathematical demonstration of this
cancellation using the BTL model under the assumption of discriminal
processes with additive biases. However, it is easy to imagine at least
two scenarios in which the zero correlation assumption is almost
certainly invalid: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics are unlikely to be equally weighted
or opposing. As a result, they may exert an uneven influence on judges'
perceptions, creating biases in their judgments. Furthermore, since the
discriminal difference of the stimuli becomes an observable outcome only
through the judges' perceptions, these biases may ultimately introduce
dependencies between the stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, thereby
favoring texts with fewer errors but weaker arguments. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, creating a statistical phenomenon known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the existence of such hierarchical structures, the
statistical handling of this additional source of dependence between
stimuli has been inadequate. For instance, when CJ data incorporates
multiple samples of stimuli from the same individuals, researchers
frequently rely on (averaged) estimated BTL scores to conduct subsequent
analyses and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

In any case, similar to Section~\ref{sec-theory-issue1a}, assuming zero
correlation between stimuli by neglecting additional relevant traits,
excluding judges based on misfit statistics, or ignoring hierarchical
(grouping) structures can cause significant statistical and measurement
issues. In general, the model may over- or underestimate how accurately
the outcome reflects the ``true'' discriminal differences between
stimuli. Such inaccuracies can result in spurious inferences about these
differences and, by extension, about the stimuli's discriminal
processes. This scenario is illustrated by Figure~\ref{fig-correlation},
when the model's discriminal difference distribution aligns with the
thick continuous line for \(\rho=0\), while the ``true'' discriminal
difference follows any discontinuous line where \(\rho \neq 0\).

In particular, neglecting relevant traits, such as judges' biases, can
cause dimensional mismatches in the BTL model, artificially inflating
the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
introducing bias into the trait's estimates \citep{Ackerman_1989}.
Excluding judges based on misfit statistics risks discarding valuable
information, which may further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially overestimating the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.

@@

In the sample-comparison model, researchers have access to a limited
number of judgments (\(n^{s}{K}\)) from a sample of judges
(\(n^{s}{J}\)), along with a specific number of texts (\(n^{s}{A}\))
from a sample of students (\(n^{s}{I}\)) drawn from the conceptual
population.

>>>>

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on the validity of CJ scores indicates their
capacity to represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Moreover, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. This study addresses these
issues by extending Thurstone's general form through a systematic and
integrated approach based on Causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key assessment design features relevant
to CJ experiments, such as the selection of judges, stimuli, and
comparisons. Section~\ref{sec-statistical} translates these theoretical
and practical elements into a probabilistic statistical model for the
analysis of dichotomous pairwise comparison data.
Section~\ref{sec-discussion} discusses the findings, limitations,
challenges and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's key insights.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Furthermore, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep[pp.~452]{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

@@

\section{Two prominent issues in CJ practice}\label{sec-theory-issues}

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
five texts, the general form would require estimating \(20\) parameters:
five modal discriminal processes, five discriminal dispersions, and
\(10\) correlations --one per comparison (see
Table~\ref{tbl-thurstone_cases}). However, a single judge could provide
only \({5 \choose 2} = 10\) unique comparisons, an insufficient data set
to estimate the required parameters.

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions into
the model. In Case I, Thurstone postulated that pairs of stimuli would
maintain a constant correlation across all comparisons. In Case II, he
allowed multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

It is evident that although Thurstone developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b} and
cautioned that its use ``should not be made without experimental test''
\citep[pp.~270]{Thurstone_1927b}, the CJ literature has nevertheless
relied the most on it in practice. This preference is particularly
notable given that Case V incorporates the most extensive set of
simplifying assumptions among all the cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}).

Furthermore, since Thurstone's primary goal was to define a
psychological scale and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
guidance on how to use these measurements when, for instance,
researchers aim to test hypotheses related to the measured traits.
Although the CJ practice has circumvented this issue by using scores'
point estimates or their transformations for such analyses, a
fundamental question remains: Is this method appropriate for addressing
these inquiries?

Thus, this section discusses these two prominent issues, as they could
undermine the reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. Specifically,
Section~\ref{sec-theory-issue1} examines the the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Moreover, Section~\ref{sec-theory-issue2} focuses on the apparent
disconnect between the approaches to trait measurement and hypothesis
testing in CJ.

\subsection{The overreliance on Case V in the statistical analysis of CJ
data}\label{sec-theory-issue1}

As previously discussed, despite relying on the most extensive set of
simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli discriminal processes, the
BTL model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). As noted by
\citet[pp.~16]{vanderLinden_et_al_2017_I}, this substitution has little
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation. Furthermore, this minimal impact is reinforced by the
fact that the Normal and Logistic distributions exhibit analogous
statistical properties, differing only by a scaling factor of
approximately \(1.7\).

However, as mentioned earlier, Thurstone originally developed Case V to
provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
He explicitly warned against its untested application, stating that its
use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}. Furthermore, he acknowledged that some
assumptions could prove problematic when researchers assess complex
traits or heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice, namely
the assumption of equal dispersion and zero correlation between stimuli.

@@

Thurstone argued that the assumption of equal dispersions may not apply
when researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
introduce judgment discrepancies due to their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of this violation
through ``misfit'' statistics. For example, labeling texts as
\emph{misfits} indicates that comparisons involving these texts result
in more judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
These discrepancies imply that the discriminal differences for
\emph{misfit} texts have broader distributions, further suggesting that
their discriminal processes exhibit more dispersion than those of other
texts. This idea can be illustrated using
Figure~\ref{fig-discriminal_process}, where text B shows a wider
discriminal dispersion than text A. As a result, the texts' discriminal
processes overlap more, making it difficult for a judge to determine
which text possesses more of the trait.

A similar line of reasoning applies to the concept of ``misfit'' judges,
whose evaluations deviate substantially from the shared consensus due to
the unique characteristics of the stimuli or the judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
These ``misfit'' judges and their associated deviations can give rise to
additional statistical and measurement issues, which we discuss in more
detail in Section~\ref{sec-theory-issue1b}.

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this idea. In this experiment, researchers can
observe the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\).
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B compared to Text A
\((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
Additionally, the figure reveals that when the discriminal dispersions
of the texts are equal \((\sigma_{B} - \sigma_{A}=0)\), the discriminal
difference is more narrow compared to situations where their dispersions
differ. Consequently, the discriminal difference is more likely to favor
Text B over Text A (shaded gray area).

@@

\caption{\label{fig-casev_issues}The effect of dispersion discrepancies
and stimuli correlation on the distribution of the discriminal
difference.}

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Consequently, the ability of the outcome to reflect \emph{true}
differences between stimuli largely depends on the validity of the
model's assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
assumption of equal dispersions. For example, when researchers correctly
assume equal dispersions between texts, the BTL model estimates a
discriminal difference distribution that accurately represents the
\emph{true} discriminal difference between them. This scenario is
illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
difference distribution aligns with the \emph{true} discriminal
difference distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
discriminal difference then ensures reliable estimates for the texts'
discriminal processes.

However, erroneously assuming equal dispersions between stimuli can give
rise to significant statistical and measurement issues. Specifically,
the model may overestimate the trait's reliability, that is, the degree
to which the outcome accurately reflects the \emph{true} discriminal
differences between stimuli. This overestimation, in turn, results in
spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
that misfit statistics highlight these critical differences in
dispersions, the conventional CJ practice of excluding stimuli based on
these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on which stimuli are excluded from the analysis.

@@

In contrast, Thurstone assumed that stimuli were uncorrelated because
judges' biases, arising from two opposing and equally weighted effects
occurring during the pairwise comparisons, canceled each other out
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
mathematical demonstration of this cancellation using the BTL model
under the assumption of discriminal processes with additive biases.
However, the assumption of zero correlation is not met in practice in at
least two scenarios. First, when intricate aspects of multidimensional,
complex traits or heterogeneous stimuli influence judges' perceptions,
generating biases that resist cancellation. Second, when an additional
hierarchical structure is relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics are unlikely to be equally weighted
or opposing. As a result, they may exert an uneven influence on judges'
perceptions, creating biases in their judgments that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could ultimately introduce dependencies
between the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While
direct evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
existence of such hierarchical structures, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

According to the theory, the correlation \(\rho\) measures how much the
judges' perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
similar thought experiment, as the one depicted in
Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
experiment now assumes that the discriminal dispersions for both texts
remain constant. Figure~\ref{fig-correlation} reveals that as the
correlation between the texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing affects the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
outcome. Furthermore, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less narrow compared to scenarios where the texts are highly
correlated. Consequently, the discriminal difference is less likely to
favor Text B over Text A (shaded gray area).

Again, in experimental practice, researchers approach this process in
reverse. They begin by observing the outcomes and use the BTL model to
estimate the stimuli' discriminal difference and discriminal processes.
Given that the BTL model assumes independent discriminal processes
across comparisons, if this assumption holds, the model estimates a
discriminal difference distribution that accurately reflects the
\emph{true} discriminal difference between the texts. This scenario is
illustrated in Figure~\ref{fig-correlation} when the discriminal
difference distribution of the model aligns with the \emph{true}
distribution, represented by the thick continuous line corresponding to
\(\rho=0\). Once more, the accuracy of the discriminal difference
ensures reliable estimates for the discriminal processes of the texts.

However, erroneously assuming zero correlation between stimuli can lead
to significant statistical and measurement issues. Specifically,
neglecting judges' biases or hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability \citep[pp.~341,
482]{Hoyle_et_al_2023}. In other words, the outcome may fail to
accurately reflect the \emph{true} discriminal differences between
stimuli. These inaccuracies can result in spurious conclusions about
these differences \citep[pp.~370]{McElreath_2020} and, by extension, the
underlying discriminal processes \citep{Ackerman_1989}. This issue is
illustrated in Figure~\ref{fig-correlation} when the model discriminal
difference distribution follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \((\rho \neq 0)\). Conversely, excluding judges
based on misfit statistics risks discarding valuable information,
potentially biasing the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Again, the direction
and magnitude of these biases are often unpredictable, as they depend on
which judges are excluded from the analysis.

@@

\subsubsection{The sample mechanism}\label{sec-theory-theoretical_SC1}
